---
title: "STA207 Homework 3"
author: "Juanjuan Hu, Zhen Zhang"
date: "January 28, 2016"
output: pdf_document
---
# 23.13
```{r, echo=FALSE,cache=FALSE, include=FALSE}
library(xlsx)
kidney <- read.xlsx2("~/Desktop/201601-03/sta207/homework/hw1/kidney.xlsx", 1)
```

```{r}
transform_function <- function(x) {
  log10(x + 1)
}
kidney$Days <- transform_function(as.numeric(as.character(kidney$Days)))
kidney <- subset(kidney, Duration != 2 | Weight != 1) # empty y_21 cell
```

## (a)
The full model is:

$Y_{ijk} = \mu_{..}+\alpha_1*X_{ij1}+\beta_1*X_{ij2}+\beta_2*X_{ij3}+\epsilon_{ijk}$

The reduced model for testing for factor A main effects is:

$Y_{ijk} = \mu_{..}+\beta_1*X_{ij2}+\beta_2*X_{ij3}+\epsilon_{ijk}$

The reduced model for testing for factor B main effects is:

$Y_{ijk} = \mu_{..}+\alpha_1*X_{ij1}+\epsilon_{ijk}$

Where:

$X_{ij1} = 
\begin{cases}
1,& \text{if case from level 1 for factor A}\\
-1,& \text{if case from level 2 for factor A}
\end{cases}$

$X_{ij2} = 
\begin{cases}
1,& \text{if case from level 1 for factor B}\\
-1,& \text{if case from level 3 for factor B}\\
0,& \text{otherwise}
\end{cases}$

$X_{ij3} = 
\begin{cases}
1,& \text{if case from level 2 for factor B}\\
-1,& \text{if case from level 3 for factor B}\\
0,& \text{otherwise}
\end{cases}$

## (b)
```{r}
kidney$X1 = ifelse(kidney$Duration == 1, 1, -1)
kidney$X2 = ifelse(kidney$Weight == 1,1, ifelse(kidney$Weight == 3, -1, 0))
kidney$X3 = ifelse(kidney$Weight == 2,1, ifelse(kidney$Weight == 3, -1, 0))
fullModel = with(kidney, lm(Days~X1+X2+X3))
fullModel
sse.full = anova(fullModel)[4,2]
sse.full
reducedModelTestA = with(kidney, lm(Days~X2+X3))
reducedModelTestA
sse.reduceA = anova(reducedModelTestA)[3,2]
sse.reduceA
reducedModelTestB = with(kidney, lm(Days~X1))
reducedModelTestB
sse.reduceB = anova(reducedModelTestB)[2,2]
sse.reduceB
```
From R output above, we can fit the full model as:

$\hat{Y} = 0.66939 +0.11733*X_{ij1}-0.34323*X_{ij2}+0.02608*X_{ij3}$

And the SSE of full model is `r sse.full`.

The reduced model for testing A main effects is:

$\hat{Y} = 0.70850-0.26502*X_{ij2}-0.01303*X_{ij3}$

The corresponding SSE is `r sse.reduceA`.

The reduced model for testing B main effects is:

$\hat{Y} = 0.75520+0.03152*X_{ij1}$

The corresponding SSE is `r sse.reduceB`.

**TESTING A MAIN EFFECTS:**

$H_0: \alpha_1 = 0$

$H_1: \alpha_1 \neq 0$

$F^*=\frac{(`r sse.reduceA`-`r sse.full`)/1}{(`r sse.full`)/46}=0.5506265/0.0976048=5.6414$

$F(0.95, 1, 46) = 4.0517$

$p-val = 0.02176$

The decision rule is: if $F^*$ is greater thatn 4.0517, then reject $H_0$, otherwise, accept $H_1$.
Here, $5.6414 \geq 4.0517$, so we reject $H_0$, concluding that factor A main effects are present. The p-value is 0.02176, which is less than 0.05, leading to the same conclusion.

**TESTING B MAIN EFFECTS:**

$H_0: \beta_1 =\beta_2= 0$

$H_1: not all \beta equal to 0$

$F^*=\frac{(`r sse.reduceB`-`r sse.full`)/2}{(`r sse.full`)/46}=1.307215/0.0976048=13.39294$

$F(0.95, 2, 46) = 3.1996$

$p-val = 2.608231e-05$

The decision rule is: if $F^*$ is greater thatn 3.1996, then reject $H_0$, otherwise, accept $H_1$.
Here, $13.39294 \geq 4.0517$, so we reject $H_0$, concluding that factor B main effects are present. The p-value is almost zero, which is less than 0.05, leading to the same conclusion.

# 23.19
## (a)
The ANOVA model is:

$Y_{ij} = \mu_{..} + \rho_i + \tau_j + \epsilon_{ij}$, Where $i = 1, 2,..., 5$, $j = 1, 2, 3$

The corresponding regression model is:

$Y_{ij} = \mu_{..}+\rho_1*X_{ij1}+\rho_2*X_{ij2}+\rho_3*X_{ij3}+\rho_4*X_{ij4}+\tau_1*X_{ij5}+\tau_2*X_{ij6}+\epsilon_{ij}$

Where:

$X_{ij1} = 
\begin{cases}
1,& \text{if case from block 1}\\
-1,& \text{if case from block 5} \\
0, & \text{otherwise}
\end{cases}$

$X_{ij2} = 
\begin{cases}
1,& \text{if case from block 2}\\
-1,& \text{if case from block 5}\\
0, & \text{otherwise}
\end{cases}$

$X_{ij3} = 
\begin{cases}
1,& \text{if case from block 3}\\
-1,& \text{if case from block 5}\\
0, & \text{otherwise}
\end{cases}$

$X_{ij4} = 
\begin{cases}
1,& \text{if case from block 4}\\
-1,& \text{if case from block 5}\\
0, & \text{otherwise}
\end{cases}$

$X_{ij5} = 
\begin{cases}
1,& \text{if case from treatment 1}\\
-1,& \text{if case from treatment 3}\\
0, & \text{otherwise}
\end{cases}$

$X_{ij6} = 
\begin{cases}
1,& \text{if case from treatment 2}\\
-1,& \text{if case from treatment 3}\\
0, & \text{otherwise}
\end{cases}$

## (b)
The reduced model for testing for differences in the mean reductions in lipid level for treatment is:

$Y_{ij} = \mu_{..}+\rho_1*X_{ij1}+\rho_2*X_{ij2}+\rho_3*X_{ij3}+\rho_4*X_{ij4}+\epsilon_{ij}$

## (c)
```{r}
Yij = c(0.73, 0.67, 0.15, 0.86, 0.75, 0.21, 0.94, 0.81, 0.26, 1.4, 1.32, 0.75, 1.62, 1.41, 0.78)
obs = data.frame(matrix(Yij, 5,3,2))
rownames(obs) = c("block1", "block2", "block3", "block4", "block5")
names(obs) = c("treatment1", "treatment2", "treatment3")
obs[1,3] = NA
obs[5,1] = NA
obs
```

```{r}
Y = c(0.73, 0.67, 0.86, 0.75, 0.21, 0.94, 0.81, 0.26, 1.4, 1.32, 0.75, 1.41, 0.78)
X1 = c(1,1,0,0,0,0,0,0,0,0,0,-1,-1)
X2 = c(0,0,1,1,1,0,0,0,0,0,0,-1,-1)
X3 = c(0,0,0,0,0,1,1,1,0,0,0,-1,-1)
X4 = c(0,0,0,0,0,0,0,0,1,1,1,-1,-1)
X5 = c(1,0,1,0,-1,1,0,-1,1,0,-1,0,-1)
X6 = c(0,1,0,1,-1,0,1,-1,0,1,-1,1,-1)
df = cbind(Y, X1, X2, X3, X4, X5,X6)
df = data.frame(df)
```

```{r}
fullModel2 = with(df, lm(Y~X1+X2+X3+X4+X5+X6))
fullModel2
sse.full2 = anova(fullModel2)[7,2]
sse.full2
reducedModel2 = with(df, lm(Y~X1+X2+X3+X4))
reducedModel2
sse.reduce2 = anova(reducedModel2)[5,2]
sse.reduce2
```

From R output above, we can fit the full model as:

$\hat{Y} = 0.8294 -0.3361*X_1-0.2227*X_2 -0.1594*X_3+0.3273*X_4+0.2508*X_5+0.1626*X_6$

And the SSE of full model is `r sse.full2`.

The reduced model for testing differences in the treatment is:

$\hat{Y} = 0.8457 -0.1457*X_1-0.2390*X_2 -0.1757*X_3+0.311*X_4$

The corresponding SSE is `r sse.reduce2`.

**TESTING TREATMENT EFFECTS:**

$H_0: \tau_1 = \tau_2 = 0$

$H_1: not all equal to zero$

$F^*=\frac{(`r sse.reduce2`-`r sse.full2`)/2}{(`r sse.full2`)/6}=0.4753/0.00058=819.48$

$F(0.95, 2, 6) = 5.1433$

The decision rule is: if $F^*$ is greater thatn 5.1433, then reject $H_0$, otherwise, accept $H_1$. Here, $819.48 \geq 4.0517$, so we reject $H_0$, concluding that the mean reductions in lipid level differ for the three diets. The result is the same as obtained in Problem 23.17d.

## (d)
```{r}
vcov(fullModel2)
```
Construct: $L = \tau_1 - \tau_3 = 2*\tau_1 + \tau_2$

$\hat{L} = 2*\hat{\tau_1} + \hat{\tau_2} = 2*0.2508+0.1626 = 0.6642$

According to the covariance matrix of model coefficients, $s^2 \{\hat{\tau_1}\} = 1.051128e-04$, 
$s^2 \{\hat{\tau_2}\} = 8.656346e-05$, $s\{\hat{\tau_1},\hat{\tau_2}\}=-4.328173e-05$. Therefore:

$s\{\hat{L}\}=sqrt(4*1.051128e-04+8.656346e-05+4*(-4.328173e-05))=0.0182726$

$t(0.99,6)=3.142668$

$\hat{L}+t(0.99,6)*s\{\hat{L}\}=0.6642+3.142668*0.0182726=0.7216247$
$\hat{L}-t(0.99,6)*s\{\hat{L}\}=0.6642-3.142668*0.0182726=0.6067753$

Therefore, the 98% confidence interval for diffrence in diet1 and diet3 is [0.6068, 0.7216]. We can find that the CI does not include zero, indicating that mean reduction in lipid for diet 1 is significantly larger than the reduction for diet3.





